---
title: "00_species-lists"
author: "Kate Sheridan"
date: "3/23/2023"
output: html_document
---

** Last updated March 20 2023 **

This script extracts species lists from each survey in GAPP for 12S and COI.

The cleaning could be done in fewer steps but as of right now more exploration might be needed at a later date.

*This is currently pre-qc.*
This needs to be updated as more data come in and the data are cleaned.

```{r setup, include=FALSE}
library(tidyverse)
library(here)

```

# 1a

Depth transect Atlantic

## 12S

Not yet


## COI

```{r}
taxonomy_1a_coi <- read.delim(here('rawdata', '1a_depth-transect_atlantic', 'COI',  'taxonomy_table.CO1.NCBI_NT+customDB.iterative_blast.LCA+best_hit.txt'), 
                           h = TRUE, fill = TRUE) %>% 
  # standardize names and remove overhanging x
  janitor::clean_names() %>%
  rename_with(., ~str_remove_all(string = .,
                 pattern = "x_")) %>%
  # column to match, eliminate any duplicates
  rename(asv = query) %>%
  distinct()


sp_1a_coi <- taxonomy_1a_coi %>%
  # no bacteria
  filter(kingdom == 'Eukaryota') %>%
  # select animal phyla; 
  # there may be some not listed here or with different backbone taxonomy check for each dataset
  filter(phylum %in% c('Acanthocephala', 'Annelida', 'Brachiopoda', 'Bryozoa',
                       'Cephalorhyncha', 'Chordata', 'Chaetognatha', 'Cnidaria', 
                       'Ctenophora', 'Dicyemida', 'Echinodermata', 'Gastrotricha', 
                       'Gnathostomulida', 'Kamptozoa', 'Micrognathozoa', 'Mollusca', 
                       'Nematoda', 'Nematomorpha', 'Nemertea', 'Orthonectida', 
                       'Phoronida', 'Placozoa', 'Platyhelminthes', 'Porifera', 'Rotifera', 
                       'Sipuncula', 'Tardigrada', 'Xenacoelomorpha')) %>%
  # only sp level ids for this, comment out for full list
  filter(lca_rank == 'species') %>%
  # remove known terrestrial and non-invert
  filter(!(class %in% c('Actinopteri', 'Elasmobranchii',
                      'Holocephali', 'Leptocardii', 'Myxini',
                      'Petromyzonti', 'Mammalia', 'Insecta'))) %>%
  # select only species column
  select(species) %>%
  distinct() %>%
  # clean up column
  mutate(verbatim_species = species) %>%
  # these may not be comprehensive; see taxonomy key script for comprehensive list
  # first remove special characters that would be near the end
  mutate(species = str_remove_all(species, "[[:digit:]]")) %>%
  mutate(species = str_remove_all(species, "-")) %>%
  # remove end of line whitespace
  mutate(species = trimws(species)) %>%
  # remove four characters worth of capital letters from the end
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{2})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  # trim and four more (in case of 'words')
  mutate(species = trimws(species)) %>%
  # remove four characters worth of capital letters from the end
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{2})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = trimws(species)) %>%
  #remove sp. from the end
  mutate(species = str_remove_all(species, "(sp\\.)$")) %>%
  mutate(species = trimws(species)) %>%
  # remove verbatim unless needed
  select(species) %>%
  # some entries will be duplicate now
  distinct() %>%
  # only species epithets, though we may want a generic list sometime
  filter(str_detect(species, ' ') == TRUE)
  
  
write_csv(sp_1a_coi, here('processeddata', 'splists', '20230323_gapp-1a-atlantic_coi-sp.csv'))


```



# 1b

depth transect vector pacific

## 12S


```{r 12S-tully}
taxonomy_1b_12s <- read.delim(here('rawdata', '1b_depth-transect_vector-pacific', '12S',  'taxonomy_table.12S.NCBI_NT.96sim.LCA+besthit.txt'), 
                           h = TRUE, fill = TRUE) %>% 
  # standardize names and remove overhanging x
  janitor::clean_names() %>%
  rename_with(., ~str_remove_all(string = .,
                 pattern = "x_")) %>%
  # column to match, eliminate any duplicates
  rename(asv = query) %>%
  distinct()


sp_1b_12s <- taxonomy_1b_12s %>%
  # fish only; check if more than Actinopteri
  filter(class %in% c('Actinopteri', 'Chondrichthyes')) %>% 
  # only sp level ids for this, comment out for full list
  filter(lca_rank == 'species') %>%
  select(species) %>%
  distinct()
# there may be clean up work here, maybe not.



write_csv(sp_1b_12s, here('processeddata', 'splists', '20230323_gapp-1b-pacific_12sfish-sp.csv'))

```


## COI

```{r}
taxonomy_1b_coi <- read.delim(here('rawdata', '1b_depth-transect_vector-pacific', 'COI',  'taxonomy_table.CO1.NCBI_NT+customDB.iterative_blast.txt'), 
                           h = TRUE, fill = TRUE) %>% 
  # standardize names and remove overhanging x
  janitor::clean_names() %>%
  rename_with(., ~str_remove_all(string = .,
                 pattern = "x_")) %>%
  # column to match, eliminate any duplicates
  rename(asv = query) %>%
  distinct()


sp_1b_coi <- taxonomy_1b_coi %>%
  # no bacteria
  filter(kingdom == 'Eukaryota') %>%
  # select animal phyla; 
  # there may be some not listed here or with different backbone taxonomy check for each dataset
  filter(phylum %in% c('Acanthocephala', 'Annelida', 'Brachiopoda', 'Bryozoa',
                       'Cephalorhyncha', 'Chordata', 'Chaetognatha', 'Cnidaria', 
                       'Ctenophora', 'Dicyemida', 'Echinodermata', 'Gastrotricha', 
                       'Gnathostomulida', 'Kamptozoa', 'Micrognathozoa', 'Mollusca', 
                       'Nematoda', 'Nematomorpha', 'Nemertea', 'Orthonectida', 
                       'Phoronida', 'Placozoa', 'Platyhelminthes', 'Porifera', 'Rotifera', 
                       'Sipuncula', 'Tardigrada', 'Xenacoelomorpha')) %>%
  # only sp level ids for this, comment out for full list
  filter(lca_rank == 'species') %>%
  # remove known terrestrial and non-invert
  filter(!(class %in% c('Actinopteri', 'Elasmobranchii',
                      'Holocephali', 'Leptocardii', 'Myxini',
                      'Petromyzonti', 'Mammalia', 'Insecta'))) %>%
  # select only species column
  select(species) %>%
  distinct() %>%
  # clean up column
  mutate(verbatim_species = species) %>%
  # these may not be comprehensive; see taxonomy key script for comprehensive list
  # first remove special characters that would be near the end
  mutate(species = str_remove_all(species, "[[:digit:]]")) %>%
  mutate(species = str_remove_all(species, "-")) %>%
  mutate(species = str_remove_all(species, "(cf\\. )")) %>%
  mutate(species = str_remove_all(species, "(complex )")) %>%
  # remove end of line whitespace
  mutate(species = trimws(species)) %>%
  # remove four characters worth of capital letters from the end
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{2})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  # trim and four more (in case of 'words')
  mutate(species = trimws(species)) %>%
  # remove four characters worth of capital letters from the end
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{2})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = trimws(species)) %>%
  #remove sp. from the end
  mutate(species = str_remove_all(species, "(sp\\.)$")) %>%
  mutate(species = trimws(species)) %>%
  # remove verbatim unless needed
  select(species) %>%
  # some entries will be duplicate now
  distinct() %>%
  # only species epithets, though we may want a generic list sometime
  filter(str_detect(species, ' ') == TRUE)
  
  
write_csv(sp_1b_coi, here('processeddata', 'splists', '20230323_gapp-1b-pacific_coi-sp.csv'))


```



# 2 Calvert

## 12S

not yet


## COI


```{r}
taxonomy_2_coi <- read.delim(here('rawdata', '2_spatial-survey_calvert-island', 'COI',  'taxonomy_table.CO1.NCBI_NT+customDB.iterative_blast.txt'), 
                           h = TRUE, fill = TRUE) %>% 
  # standardize names and remove overhanging x
  janitor::clean_names() %>%
  rename_with(., ~str_remove_all(string = .,
                 pattern = "x_")) %>%
  # column to match, eliminate any duplicates
  rename(asv = query) %>%
  distinct()


sp_2_coi <- taxonomy_2_coi %>%
  # no bacteria
  filter(kingdom == 'Eukaryota') %>%
  # select animal phyla; 
  # there may be some not listed here or with different backbone taxonomy check for each dataset
  filter(phylum %in% c('Acanthocephala', 'Annelida', 'Brachiopoda', 'Bryozoa',
                       'Cephalorhyncha', 'Chordata', 'Chaetognatha', 'Cnidaria', 
                       'Ctenophora', 'Dicyemida', 'Echinodermata', 'Gastrotricha', 
                       'Gnathostomulida', 'Kamptozoa', 'Micrognathozoa', 'Mollusca', 
                       'Nematoda', 'Nematomorpha', 'Nemertea', 'Orthonectida', 
                       'Phoronida', 'Placozoa', 'Platyhelminthes', 'Porifera', 'Rotifera', 
                       'Sipuncula', 'Tardigrada', 'Xenacoelomorpha')) %>%
  # only sp level ids for this, comment out for full list
  filter(lca_rank == 'species') %>%
  # remove known terrestrial and non-invert
  filter(!(class %in% c('Actinopteri', 'Elasmobranchii',
                      'Holocephali', 'Leptocardii', 'Myxini',
                      'Petromyzonti', 'Mammalia', 'Insecta'))) %>%
  # select only species column
  select(species) %>%
  distinct() %>%
  # clean up column
  mutate(verbatim_species = species) %>%
  # these may not be comprehensive; see taxonomy key script for comprehensive list
  # first remove special characters that would be near the end
  mutate(species = str_remove_all(species, "[[:digit:]]")) %>%
  mutate(species = str_remove_all(species, "-")) %>%
  mutate(species = str_remove_all(species, "_")) %>%
  mutate(species = str_remove_all(species, "(cf\\. )")) %>%
  mutate(species = str_remove_all(species, "(aff\\. )")) %>%
  mutate(species = str_remove_all(species, "(gen\\. )")) %>%
  mutate(species = str_remove_all(species, "(complex )")) %>%
  # remove end of line whitespace
  mutate(species = trimws(species)) %>%
  # remove four characters worth of capital letters from the end
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{2})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  # trim and four more (in case of 'words')
  mutate(species = trimws(species)) %>%
  # remove four characters worth of capital letters from the end
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{2})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = trimws(species)) %>%
  #remove sp. from the end
  mutate(species = str_remove_all(species, "(sp\\.)$")) %>%
  mutate(species = str_remove_all(species, "(sp\\. )")) %>%
  mutate(species = trimws(species)) %>%
  # remove verbatim unless needed
  select(species) %>%
  # some entries will be duplicate now
  distinct() %>%
  # only species epithets, though we may want a generic list sometime
  filter(str_detect(species, ' ') == TRUE)
  
  
write_csv(sp_2_coi, here('processeddata', 'splists', '20230323_gapp-2-calvert_coi-sp.csv'))


```


# 3 Tully

## 12S


```{r 12S-tully}
taxonomy_3_12s <- read.delim(here('rawdata', '3_deep-water_tully', '12S',  'taxonomy_table.12S.NCBI_NT.96sim.txt'), 
                           h = TRUE, fill = TRUE) %>% 
  # standardize names and remove overhanging x
  janitor::clean_names() %>%
  rename_with(., ~str_remove_all(string = .,
                 pattern = "x_")) %>%
  # column to match, eliminate any duplicates
  rename(asv = query) %>%
  distinct()


sp_3_12s <- taxonomy_3_12s %>%
  # fish only; check if more 
  filter(class %in% c('Actinopteri', 'Chondrichthyes')) %>% 
  # only sp level ids for this, comment out for full list
  filter(lca_rank == 'species') %>%
  select(species) %>%
  distinct()
# there may be clean up work here, maybe not.


unique(taxonomy_3_12s$class)


write_csv(sp_3_12s, here('processeddata', 'splists', '20230323_gapp-3-tully_12sfish-sp.csv'))

```



## COI


```{r coi-tully}
taxonomy_3_coi <- read.delim(here('rawdata', '3_deep-water_tully', 'COI',  'taxonomy_table.CO1.NCBI_NT+customDB.iterative_blast.txt'), 
                           h = TRUE, fill = TRUE) %>% 
  # standardize names and remove overhanging x
  janitor::clean_names() %>%
  rename_with(., ~str_remove_all(string = .,
                 pattern = "x_")) %>%
  # column to match, eliminate any duplicates
  rename(asv = query) %>%
  distinct()


sp_3_coi <- taxonomy_3_coi %>%
  # no bacteria
  filter(kingdom == 'Eukaryota') %>%
  # select animal phyla; 
  # there may be some not listed here or with different backbone taxonomy check for each dataset
  filter(phylum %in% c('Acanthocephala', 'Annelida', 'Brachiopoda', 'Bryozoa',
                       'Cephalorhyncha', 'Chordata', 'Chaetognatha', 'Cnidaria', 
                       'Ctenophora', 'Dicyemida', 'Echinodermata', 'Gastrotricha', 
                       'Gnathostomulida', 'Kamptozoa', 'Micrognathozoa', 'Mollusca', 
                       'Nematoda', 'Nematomorpha', 'Nemertea', 'Orthonectida', 
                       'Phoronida', 'Placozoa', 'Platyhelminthes', 'Porifera', 'Rotifera', 
                       'Sipuncula', 'Tardigrada', 'Xenacoelomorpha')) %>%
  # only sp level ids for this, comment out for full list
  filter(lca_rank == 'species') %>%
  # remove known terrestrial and non-invert
  filter(!(class %in% c('Actinopteri', 'Elasmobranchii',
                      'Holocephali', 'Leptocardii', 'Myxini',
                      'Petromyzonti', 'Mammalia', 'Insecta'))) %>%
  # select only species column
  select(species) %>%
  distinct() %>%
  # clean up column
  mutate(verbatim_species = species) %>%
  # these may not be comprehensive; see taxonomy key script for comprehensive list
  # first remove special characters that would be near the end
  mutate(species = str_remove_all(species, "[[:digit:]]")) %>%
  mutate(species = str_remove_all(species, "-")) %>%
  mutate(species = str_remove_all(species, "_")) %>%
  mutate(species = str_remove_all(species, "(cf\\. )")) %>%
  mutate(species = str_remove_all(species, "(aff\\. )")) %>%
  mutate(species = str_remove_all(species, "(gen\\. )")) %>%
  mutate(species = str_remove_all(species, "(complex )")) %>%
  # remove end of line whitespace
  mutate(species = trimws(species)) %>%
  # remove four characters worth of capital letters from the end
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{2})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  # trim and four more (in case of 'words')
  mutate(species = trimws(species)) %>%
  # remove four characters worth of capital letters from the end
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{2})$")) %>%
  mutate(species = str_remove_all(species, "([A-Z]{1})$")) %>%
  mutate(species = trimws(species)) %>%
  # hanging single lowercase
  mutate(species = str_remove_all(species, "( [a-z]{1})$")) %>%
  mutate(species = trimws(species)) %>%
  # remove sp. from the end
  mutate(species = str_remove_all(species, "(sp\\.)$")) %>%
  mutate(species = str_remove_all(species, "(sp\\. )")) %>%
  mutate(species = trimws(species)) %>%
  # remove verbatim unless needed
  select(species) %>%
  # some entries will be duplicate now
  distinct() %>%
  # only species epithets, though we may want a generic list sometime
  filter(str_detect(species, ' ') == TRUE)
  
  
write_csv(sp_3_coi, here('processeddata', 'splists', '20230323_gapp-3-tully_coi-sp.csv'))


```