---
title: "taxonomy_workflow_chunks"
output: html_document
date: "2023-04-11"
---

Copypasta from previous scripts




Extracting fresh taxonomy for everyone, including freshwater-marine status.

This function pulls from worms using worrms
```{r worms-function}
library(worrms)

## requires worrms to be loaded
# input is vector of species names
# use wm_records_names to extract 
# this doesn't need the + signs

# right now this breaks if not-found
search_records_worms <- function(spnames) {
  search <- tibble()
  for (i in spnames) {
    print(paste0('searching for ', i))
    record <- wm_records_names(i, marine_only = FALSE)
    message('done')
    search <- append(search, record)
  }
  names(search) <- spnames
  search_output <- map_dfr(.x = search, ~ data.frame(.x), .id = 'query') %>%
    janitor::clean_names() %>%
    select(!(c(url, taxon_rank_id, citation, lsid, modified)))
  return(search_output)
}
```



I make a dataframe with species to search since often its from something else, there's an issue with the worms thing so i have an add plus sign section in the beginning


```{r bc-taxonomy}
# check which ones are in the database
#then replace resolved
fish_search <- fish_list %>% 
  select(c(species)) %>%
  ## temporary issue requires replacing whitespace with '+'
  mutate(worms_query = gsub(" ", "+", species)) %>%
  distinct() %>%
  rename(query = species)


# 2: worms
fish_worms <- get_wormsid_(sci_com = fish_search$worms_query, marine_only = FALSE,
                           accepted = FALSE)
# get_wormsid_ gives back null and length 0 elements
## first remove with compact and discard then map_dfr
fish_worms2 <- fish_worms %>% 
  compact() %>%
  discard( ~ nrow(.x) == 0) %>%
  map_dfr( ~ data.frame(.x), .id = 'query')


# this cleaning step is relatively temporary until they fix the curl issue
fish_worms2 <- fish_worms2 %>%
  #rename(taxa = scientificname) %>%
  left_join(fish_search, by = c('query' = 'worms_query')) %>%
  select(!(query)) %>%
  rename(taxa = query.y)

# now match species to sciname to remove extra rows
fish_worms2 <- fish_worms2 %>%
  # remove higher taxonomy with species-level hits
  filter(!(str_detect(taxa, " ") == FALSE & str_detect(scientificname, " ") == TRUE)) %>%
  filter(!(!(taxa == scientificname))) %>%
  select(!(c(status))) %>%
  rename(worms_aphiaid = AphiaID,
        worms_sciname = scientificname) %>%
  distinct()


# extract higher and info about freshwater
### this is my custom function!
fish_records <- search_records_worms(fish_worms2$worms_sciname)

# refine results to be useful
fish_records <- fish_records %>%
  # only animal hits
  filter(kingdom == 'Animalia') %>%
  # generate terrestrial-only and freshwater-only columns
  mutate(freshwater_only = ifelse((is_marine %in% c(0, NA) ==TRUE) & (is_brackish %in% c(0,NA) == TRUE) & (is_freshwater == 1),  "1", "0")) %>%
    mutate(terrestrial_only = ifelse((is_marine %in% c(0, NA) ==TRUE) & (is_brackish %in% c(0,NA) == TRUE) & (is_terrestrial == 1),  "1", "0")) %>%
  distinct() %>%
  # rename to format
  rename(worms_aphiaid = valid_aphia_id,
         original_aphiaid =aphia_id,
         worms_match = scientificname,
         worms_sciname = valid_name
         ) %>%
  # remove unneeded columns
  select(!(c(authority, parent_name_usage_id, is_freshwater, 
             is_terrestrial, is_marine, is_brackish, is_extinct, match_type))) %>%
  relocate(c(worms_sciname, worms_aphiaid), .before = original_aphiaid) %>%
  # rank lowercase
  mutate(rank = tolower(rank))


```


Then I save the output as csv and find problem species -- that weren't found in worrms or any of hte other databases I search

```{r}
fish_problems <- fish_caution %>%
  filter(!(species %in% fish_list_taxa2$species))
```

